###README: Human-Activity-Recognition-Using-Smartphones
#####This analysis performs a data cleansing process and generate tidy datasets at the end. Specifically, it follows following steps:

#####1. Merges the training and the test sets to create one data set.

```{r step1, echo=TRUE}
#acquire url and name of raw dataset
harus.url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
harus.file <- basename(harus.url)

#download dataset if it cannot be find at local drive
if(!(file.exists(harus.file))){
    download.file(harus.url,destfile=harus.file,method="auto")        
}

#obtain a list of all files containing in unziped folder
harus.file.list <- unzip(harus.file)

#read the required dataset accordingly
features <- read.table(file=harus.file.list[2])
feature.names <- features[,2]
x.test <- read.table(file=harus.file.list[15])
y.test <- read.table(file=harus.file.list[16])
subject.test <- read.table(file=harus.file.list[14])
x.train <- read.table(file=harus.file.list[27])
y.train <- read.table(file=harus.file.list[28])
subject.train <- read.table(file=harus.file.list[26])

#merge obtained dataset into one major dataset    
test.merged <- cbind(x.test,y.test,subject.test)
train.merged <- cbind(x.train,y.train,subject.train)
harus.data <- rbind(train.merged,test.merged)
colnames(harus.data)<-c(as.vector(feature.names),"y","subject")

#print some basic information re produced dataset
colna <- head(names(harus.data),n=3)
rownum <- sum(table(harus.data[,1]))
```
There are `r rownum` observations in the dataset created in step 1, and the first variables are `r colna`. 

#####2. Extracts only the measurements on the mean and standard deviation for each measurement. 
```{r step2, echo=TRUE}
#obtain the entire list of variables of dataset
harus.col <- names(harus.data)

#claim the key measurements (mean, std) we want to keep in new dataset
measure.mean <- "mean()"
measure.std <- "std()"

#identify the columns we want to keep and the columns we want to eliminate
#and implement these modifications to dataset
a <- grepl(measure.mean,harus.col,fixed=T)
b <- grepl(measure.std,harus.col,fixed=T)
c <- a | b
harus.subset <- harus.data[,c]

#print the remaining measurements in subsetted dataset
names(harus.subset)
```

#####3. Uses descriptive activity names to name the activities in the data set
```{r step3, echo=TRUE}
#add two new columns representing activity and subject
harus.subset$y <- harus.data[,562]
harus.subset$Subject <- harus.data[,563]

#remove special characters like '()' in column names to make them more readable
names(harus.subset) <- gsub("\\()","", names(harus.subset))

#load the activity_labels.txt file and make its column names be identical to the names in main dataset
act.name <- read.table(harus.file.list[1])
colnames(act.name)<-c("y","Activity")

#load plyr package to do the merging process that adds activity labels to each observation in dataset 
library(plyr)
harus.all <- arrange(join(harus.subset, act.name),y)

#print the activity distribution by subject
table(harus.all$y, harus.all$Activity)
```

#####4. Appropriately labels the data set with descriptive variable names. 
```{r step4,echo=TRUE}
#remove duplicated variable
harus.tidy <- harus.all[,-67]

#remove special characters ("-") in column names and transfer the names to lowercases
names(harus.tidy) <- gsub("-","",tolower(names(harus.tidy)))

#transfer subject and activity to factors for the convinience of further analysis
harus.tidy$subject <- as.factor(harus.tidy$subject)
harus.tidy$activity <- as.factor(harus.tidy$activity)
```

#####5. Creates a second, independent tidy data set with the average of each variable for each activity and each subject. 
```{r step5,echo=TRUE}
#remove the unnecessary std related variables in dataset
d <- grepl("std",names(harus.tidy),fixed=T)    
harus.tidy <- harus.tidy[,!d]

#use reshape package to melt the dataset with activity and subject as ids and the rest are variables
library(reshape)
harus.molten <- melt(harus.tidy, id.vars=c("activity","subject"))

#restructure the dataset into the one with the average of each variable by each activity and subject
harus.cast <- cast(subject + variable ~ activity, data = harus.molten, fun = mean)

#print the first 5 rows of constructed dataset
head(harus.cast,n=5)

#output the tidy main dataset and tidy average dataset
write.table(harus.tidy, "./tidy_dataset.csv", sep=",")
write.table(harus.cast, "./tidy_dataset_avg.csv", sep=",")
```
Thank you for reading!
Ivan Liu
